---
title: "Final Project: Asthma and Mortality"
author: "Whitney Crook"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```
# ABSTRACT

Write at the end. Summarize the study.

# BACKGROUND

Asthma is a chronic lung condition that can cause the airways to become inflamed at times making it harder to breathe. It is a very common condition in the United States. According to the CDC, about 1 in 13 people have asthma in the United States (CDC, 2025). Asthma symptoms can be triggered by many things, including pollen, exercise, cold air, or air pollution.

Air pollution can increase the risk of developing asthma and as well as making symptoms worse if a person already has asthma. The airborne particles that causes air pollution can irritate airways which triggers asthma symptoms (Asthma and Allergy Foundation of America). 

There is typically more air pollution in more highly populated regions. For this reason, we hypothesize that more urbanized regions with higher population density will be associated with higher asthma prevalence. The Northeastern United States are more highly populated than the West, so we predict that the Northeast will have a higher asthma prevalence.

```{r data for map, include=FALSE}

asthma_data <- data.frame(
  state = c("Alabama","Alaska","Arizona","Arkansas","California","Colorado","Connecticut",
            "Delaware","District of Columbia","Florida","Georgia","Hawaii","Idaho","Illinois",
            "Indiana","Iowa","Kansas","Kentucky","Louisiana","Maine","Maryland","Massachusetts",
            "Michigan","Minnesota","Mississippi","Missouri","Montana","Nebraska","Nevada",
            "New Hampshire","New Jersey","New Mexico","New York","North Carolina","North Dakota",
            "Ohio","Oklahoma","Oregon","Pennsylvania","Rhode Island","South Carolina",
            "South Dakota","Tennessee","Texas","Utah","Vermont","Virginia","Washington",
            "West Virginia","Wisconsin","Wyoming"),
  
  state_abrv = c("AL","AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC", "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY"),
  
  number = c(378702,58742,555997,243032,2629546,500090,356651,
             78980,60410,1646583,802783,103228,161970,856488,
             574447,238689,237705,377365,351940,146031,499134,633126,
             935666,434473,211055,497216,102617,121412,251193,
             148330,644536,171212,1611027,764909,61944,
             1042354,373879,389831,1033741,117252,368883,
             56630,636322,1760654,266105,68112,671051,667228,
             181395,503318,46973),
  
  prevalence_raw = c("9.6 (0.59)","10.7 (0.58)","9.7 (0.49)","10.5 (0.57)","8.7 (0.39)","10.8 (0.43)","12.4 (0.51)",
                     "9.9 (0.64)","11.1 (0.77)","9.3 (0.53)","9.6 (0.48)","9.1 (0.45)","11.1 (0.51)","8.7 (0.58)",
                     "11.0 (0.39)","9.7 (0.42)","10.7 (0.41)","10.8 (0.68)","10.0 (0.55)","13.1 (0.52)","10.4 (0.38)","11.3 (0.43)",
                     "11.9 (0.44)","9.9 (0.32)","9.4 (0.63)","10.4 (0.49)","11.7 (0.51)","8.1 (0.46)","10.1 (0.79)",
                     "13.1 (0.64)","8.9 (0.43)","10.4 (0.62)","10.3 (0.34)","9.2 (0.62)","10.4 (0.60)",
                     "11.4 (0.38)","12.3 (0.54)","11.5 (0.51)","10.1 (0.67)","13.3 (0.64)","9.0 (0.40)",
                     "8.3 (0.75)","11.7 (0.61)","7.9 (0.37)","11.0 (0.43)","12.9 (0.53)","9.9 (0.46)","10.9 (0.26)",
                     "12.9 (0.61)","10.9 (0.42)","10.5 (0.65)"), 
  
  population = c(5108468, 733406, 7431344, 3067732, 38965193, 5877610, 3617176, 1031890, 678972,
  22610726, 11029227, 1435138, 1964726, 12549689, 6862199, 3207004, 2940546,
  4526154, 4573749, 1395722, 6180253, 7001399, 10037261, 5737915, 2939690,
  6196156, 1132812, 1978379, 3194176, 1402054, 9290842, 2114371, 19571216,
  10835491, 783926, 11785935, 4053824, 4233358, 12961683, 1095962, 5373555,
  919318, 7126489, 30503301, 3417735, 647464, 8715698, 7812880, 1770071,
  5910955, 584057)
)

library(dplyr)
library(stringr)

asthma_data <- asthma_data %>%
  mutate(
    prevalence = as.numeric(str_extract(prevalence_raw, "^[0-9.]+")),
    SE = as.numeric(str_extract(prevalence_raw, "(?<=\\().+?(?=\\))")),
    state_lower = tolower(state),
    region = state.region[match(state, state.name)]
  )

asthma_data <- na.omit(asthma_data)

```

```{r prevalence by state, include=TRUE, eval=TRUE, warning=FALSE, fig.align='center', fig.width=6, fig.height=3}

library(ggplot2)
library(maps)
library(dplyr)

# Map data
states_map <- map_data("state")

# Merge map data with asthma data
map_data <- states_map %>%
  left_join(asthma_data, by = c("region" = "state_lower"))

ggplot(map_data, aes(long, lat, group = group, fill = prevalence)) +
  geom_polygon(color = "gray90", size = 0.3) +  # state borders
  geom_polygon(aes(fill = prevalence), color = "white") +
  coord_fixed(1.3) +
 scale_fill_gradientn(
  name = "Asthma Prevalence (%)",
  colors = c("yellow2", "orange", "red", "red4")
) +
  labs(
    title = "Asthma Prevalence by U.S. State",
    subtitle = "Higher prevalence shown in darker colors"
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )

```

```{r region map, include=FALSE, eval=TRUE, warning=FALSE, fig.align='center', fig.width=6, fig.height=3}

# U.S. states map data
states_map <- map_data("state")

# Put the states into regions in a dataframe
state_regions <- data.frame(
  state = tolower(state.name),
  region = state.region
)

# Merge map data with region info
map_data <- states_map %>%
  left_join(state_regions, by = c("region" = "state"))

# Plot
ggplot(map_data, aes(x = long, y = lat, group = group, fill = region.y)) +
  geom_polygon(color = "white") +
  coord_fixed(1.3) +
  labs(title = "Four U.S. Census Regions",
       fill = "Region") +
  theme_void()
```

# STUDY QUESTION and HYPOTHESIS

## Question  

Do asthma prevalence and asthma mortality rates differ across U.S. regions?

## Hypothesis 

Regions with higher population density and greater urbanization will be associated with higher asthma prevalence and mortality rates.
  
## Prediction

We predict that the Northeastern United States will have a higher prevalence of asthma and asthma-related deaths because of the higher population density.

# METHODS

### Procedure



## Bar Plot Showing Asthma Prevalence by State
```{r bar plot per state, fig.align='center', fig.width = 6, fig.height = 7}

library(ggplot2)

ggplot(asthma_data, aes(x = reorder(state, prevalence), y = prevalence)) +
  geom_col(fill = "darkgray") +
  coord_flip() +
  labs(title = "Asthma Prevalence by U.S. State",
       x = "State",
       y = "Asthma Prevalence (%)") +
  theme_minimal()
```

## Scatter Plot by State
```{r scatter plot, include=TRUE, eval=TRUE, warning=FALSE, fig.align='center'}

ggplot(asthma_data, aes(x = population, y = prevalence)) +
  geom_point(aes(color=region)) +
  geom_text(aes(label = state_abrv), vjust = -0.5, size = 3)+
  labs(title = "Number of Individuals With Asthma vs. Asthma Prevalence by State",
       x = "Population",
       y = "Asthma Prevalence (%)") +
  theme_minimal()
```

## Box Plot for Asthma Prevalence by Region

To create this bar plot, we split the United States into four regions-Northeast, South, North Central, and West. The plot shows the average asthma prevalence (percentage) for each of the four regions showing the amount of error. The standard error comes from finding the average asthma percentage among all of the states included in each region. 

```{r box plot, include=TRUE, eval=TRUE, warning=FALSE, fig.align='center', fig.width=4.75, fig.height=6}

library(ggplot2)

ggplot(asthma_data, aes(x = region, y = prevalence)) +
  geom_boxplot() +
  labs(title = "Asthma Prevalence by U.S. Region",
       x = "Region",
       y = "Asthma Prevalence (%)") +
  theme_minimal()

```

# DISCUSSION

## Interpretation - Figure 1 ()



## Interpretation - Figures 2 (Bar Plot)

This bar plot supports our prediction that the Northeast would have the highest asthma prevalence

# CONCLUSION



# REFERENCES     

1. ChatGPT. OpenAI, version Jan 2025. Used as a reference for functions
    such as ggplot(), creating the map and troubleshooting error messages. Accessed 2025-11-10.
2. Centers for Disease Control and Prevention. (2024, November 21). Most recent asthma data. U.S. Department of Health and Human Services. https://www.cdc.gov/asthma-data/about/most-recent-asthma-data.html
3. National Heart, Lung, and Blood Institute. (2025). Asthma. U.S. Department of Health and Human Services. https://www.nhlbi.nih.gov/health/asthma
4. Asthma and Allergy Foundation of America. (n.d.). Air pollution and asthma. Asthma & Allergy Foundation of America. https://aafa.org/asthma/asthma-triggers-causes/air-pollution-smog-asthma/






